---
title: "Homework 4"
author: "Yue Cao"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
    toc: true
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{The ridge regression vignette}
-->

##3
Consider the sparse matrix implementation from class and the sparse add
function:
```{r}
sparse_add <- function(a, b) {
  c <- merge(a, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
  c$x <- c$x1 + c$x2
  c[, c("i", "j", "x")]
}

a <- data.frame(i = c(1, 2), j = c(1, 1), x = c(3, 1))
b <- data.frame(i = c(1, 2, 2), j = c(1, 1, 2), x = c(4.4, 1.2, 3))
sparse_add(a, b)
```

        - Implement a `sparse_multiply` function that multiplies two sparse matrices.
        - Create a new class `sparse.matrix` that has add `+`, multiply `%*%`, and transpose `t()` methods.
        - Add test-sparse-matrix.r to the testthat directory of your bis557 package to show it works.

First, implement a 'sparse_multiply' function that multiplies two sparse matrices. Here, I also implement a 'sparse_transpose' function to transpose a matrix. In this way, we can simplify the algorithm of the sparse matrix multiplication.
```{r}
library(dplyr)

sparse_transpose <- function(a) {
  at <- a
  at$i <- a$j
  at$j <- a$i
  at$x <- a$x
  at <- at[order(at$i, at$j),]
  at[, c("i", "j", "x")]
}

sparse_multiply <- function(a, b) {
  bt <- sparse_transpose(b)
  cc <- merge(a, bt, by = c("j"), all = TRUE, suffixes = c("1", "2"))
  cc$x <- cc$x1 * cc$x2
  cc <- na.omit(cc)
  cc <- cc[, c("i1", "i2", "x")]
  c <- cc %>%
    group_by(i1, i2) %>%
    summarise(x=sum(x))
  #c <- aggregate(cc$x, by=list(cc$i1, cc$i2), FUN=sum)
  names(c) <- c("i", "j", "x")
  c <- c[order(c$i, c$j),]
  c[, c("i", "j", "x")]
}

a <- data.frame(i = c(1, 2), j = c(1, 1), x = c(3, 1))
b <- data.frame(i = c(1, 2, 2), j = c(1, 1, 2), x = c(4.4, 1.2, 3))
sparse_multiply(a, b)
```

Then, we are gonna create a new class `sparse.matrix` that has add `+`, multiply `%*%`, and transpose `t()` methods as follows:
```{r}
sparse.matrix <- function(i, j, x, dims) {
  if (missing(dims)) {
    a <- data.frame(i = i, j = j , x = x)
    attr(a, "dims") <- c(max(i), max(j))
  } 
  else {
    a<- data.frame(i = i, j = j, x = x)
    attr(a, "dims") <- dims
  }
  class(a) <- c("sparse.matrix", class(a))
  a
}


'+.sparse.matrix' <- function(a, b) {
  if (attr(a, "dims")[1] != attr(b, "dims")[1] || attr(a, "dims")[2] != attr(b, "dims")[2]) {
    stop("should not add these two matrices.")
  }
  c <- merge(a, b, by = c("i", "j"), all = TRUE, suffixes = c("1", "2"))
  c$x1[is.na(c$x1)] <- 0
  c$x2[is.na(c$x2)] <- 0
  c$x <- c$x1 + c$x2
  c <- c[order(c$j),]
  attr(c, "dims") <- attr(a, "dims")
  sparse.matrix(c$i, c$j, c$x, attr(c, "dims"))
}

't.sparse.matrix' <- function(a) {
  at <- a
  at$i <- a$j
  at$j <- a$i
  at$x <- a$x
  at <- at[order(at$i, at$j),]
  attr(at, "dims")[1] <- attr(a, "dims")[2]
  attr(at, "dims")[2] <- attr(a, "dims")[1]
  sparse.matrix(at$i, at$j, at$x, attr(at, "dims")) 
}

'%*%' <- function(a, b) {
  UseMethod("%*%", a)
}

'%*%.default' <- function(a, b) {
  warning("No method to dispatch to, using default.")
  a %*% b
}

'%*%.matrix' <- function(a, b) {
  if (!inherits(b, "matrix") && !inherits(b, "Matrix")) {
    stop("b argument is not a matrix type.")
  }
  a %*% b
}

'%*%.numeric' <- function(a, b) {
  if (!inherits(b, "numeric")) {
    stop("b argument is not a numeric vector type.")
  }
  a %*% b
}

'%*%.sparse.matrix' <- function(a, b) {
  if (attr(a, "dims")[2] != attr(b, "dims")[1]) {
    stop("should not multiply these two matrices.")
  }
  cc <- merge(a, t(b), by = c("j"), all = TRUE, suffixes = c("1", "2"))
  cc$x <- cc$x1 * cc$x2
  cc <- na.omit(cc)
  cc <- cc[, c("i1", "i2", "x")]
  #c <- aggregate(cc$x, by=list(cc$i1, cc$i2), FUN=sum)
  c <- cc %>%
    group_by(i1, i2) %>%
    summarise(x=sum(x))
  names(c) <- c("i", "j", "x")
  c <- c[order(c$j),]
  attr(c, "dims")[1] <- attr(a, "dims")[1]
  attr(c, "dims")[2] <- attr(b, "dims")[2]
  sparse.matrix(c$i, c$j, c$x, attr(c, "dims"))
}
```